---
title: "First Project"
author: "My Le"
date: "2024-06-15"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
library(readr)
library(knitr)
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)
```

## 1. Shooting by Timeframes
```{r nypd, message = FALSE}
# Retrieve the data
nypd <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")

# Pick the columns to remove
columns_to_remove <- c(1,5,8,9,17:21)

# Remove the unncessary columns
nypd <- nypd[, -columns_to_remove]

# Create a new data frame to analyze time
nypd_time <- nypd[, "OCCUR_TIME"]

# Define the time frames
nypd_time <- nypd_time %>%
  mutate(TIME_FRAME = case_when(
    OCCUR_TIME >= hms::as_hms("00:00:00") & OCCUR_TIME <= hms::as_hms("06:00:00") ~ "Dawn (0AM - 6AM)",
    OCCUR_TIME >= hms::as_hms("06:01:00") & OCCUR_TIME <= hms::as_hms("12:00:00") ~ "Morning (6AM - 12PM)",
    OCCUR_TIME >= hms::as_hms("12:01:00") & OCCUR_TIME <= hms::as_hms("18:00:00") ~ "Afternoon (12PM - 6PM)",
    OCCUR_TIME >= hms::as_hms("18:01:00") & OCCUR_TIME <= hms::as_hms("23:59:59") ~ "Evening (6PM - 12AM)"
  ))

# Count the occurrences within each time frame
cases_by_time <- nypd_time %>%
  group_by(TIME_FRAME) %>%
  summarise(REPORTED_CASES = n()) %>%
  arrange(desc(REPORTED_CASES))

# Calculate the total number of reported cases
total_cases <- sum(cases_by_time$REPORTED_CASES)

# Calculate the percentage for each time frame
cases_by_time <- cases_by_time %>%
  mutate(PERCENTAGE = (REPORTED_CASES / total_cases) * 100)

# Round the percentages to 2 decimal places
cases_by_time$PERCENTAGE <- round(cases_by_time$PERCENTAGE, 2)

# Pie chart
pie(cases_by_time$PERCENTAGE, labels = cases_by_time$TIME_FRAME, main = "Percentage of Reported Cases by Time Frame")
```

### Analysis of Shooting Timeframes

**Observations**

The data suggests that a higher percentage of police shootings occur during dawn (0 AM - 6 AM) and evening (6 PM - 12 AM) hours. This coincides with periods of lower visibility, potentially leading to:

- Reduced public presence: Fewer bystanders might be present to witness or intervene in an escalating situation.
- Delayed assistance: Victims or bystanders might be less likely to seek or receive immediate help due to reduced visibility or fewer people around.

**Actionable Insights**

Based on this observation, law enforcement can consider the following actions:

- Enhanced Nightly Patrols: Strategic deployment of additional patrols in high-risk areas during nighttime hours might deter crime and increase officer presence as a deterrent.
- Improved Nighttime Visibility: Investigate the feasibility of improved street lighting or partnerships with local businesses to enhance nighttime visibility in critical areas.
- Community Outreach Programs: Strengthening relationships with community members, particularly during evenings, can foster trust and encourage reporting of suspicious activity.
- De-escalation Training with Nighttime Considerations: Tailoring de-escalation training programs to address specific challenges and potential visibility limitations encountered during nighttime scenarios.

**Drawbacks and Considerations**

- Data Limitations: This analysis is based on a limited time frame. A longer period and access to additional data (e.g., call types, locations, demographics) could provide a more nuanced understanding.
- Reporting Bias: The data might be skewed if reporting rates differ across time frames. Investigate if there are any known reporting trends based on time of day.
- Underlying Factors: While reduced visibility might play a role, it's likely not the sole factor. Other potential factors like the nature of calls for service or suspect demographics might also contribute.

## 2. Shooting by Season
```{r nypd1, message = FALSE}
# Convert OCCUR_DATE to Date format with explicit "MM/DD/YYYY" format
nypd$OCCUR_DATE <- as.Date(nypd$OCCUR_DATE, format = "%m/%d/%Y")

# Extract month from OCCUR_DATE and aggregate by month and TIME_FRAME
nypd_date <- nypd_time %>%
  mutate(MONTH = format(nypd$OCCUR_DATE, "%b")) %>%
  group_by(MONTH, TIME_FRAME) %>%
  summarise(REPORTED_SHOOTINGS = n())

# Pivot the data to have TIME_FRAME as columns and MONTH as rows
nypd_date <- nypd_date %>%
  pivot_wider(names_from = TIME_FRAME, values_from = REPORTED_SHOOTINGS, values_fill = 0) %>%
  mutate(MONTH = factor(MONTH, levels = month.abb)) %>%
  arrange(MONTH)

# Arrange columns by Dawn, Morning, Afternoon, and Evening
nypd_date <- nypd_date %>%
  select(MONTH, "Dawn (0AM - 6AM)", "Morning (6AM - 12PM)", "Afternoon (12PM - 6PM)", "Evening (6PM - 12AM)")

# Reshape data for line plot
nypd_date_long <- pivot_longer(nypd_date, cols = -MONTH, names_to = "TIME_FRAME", values_to = "REPORTED_SHOOTINGS")

# Line plot
ggplot(nypd_date_long, aes(x = MONTH, y = REPORTED_SHOOTINGS, color = TIME_FRAME, group = TIME_FRAME)) +
  geom_line() +
  labs(title = "Monthly Trends of Shootings by Time Frame",
       x = "Month",
       y = "Number of Shootings",
       color = "Time Frame") +
  scale_x_discrete(labels = month.abb) +
  theme_minimal() +
  theme(legend.position = "right")
```


### Analysis of Shooting by Season

**Observations**

The data suggests a potential link between shooting cases in New York and summer months (June-August). Here are some possible explanations for this observation:

- Increased Activity: Warmer weather might lead to more people being outdoors, potentially increasing opportunities for conflict and gun violence.
- Behavioral Changes: Hotter weather might contribute to heightened emotional states, potentially leading to more volatile situations.
- Social Gatherings: Increased social gatherings during summer might create environments where tensions can easily escalate, particularly if firearms are present.

**Actionable Insights**

Based on this observation, various entities can consider the following actions:

- Community Outreach Programs: Strengthening community engagement initiatives during summer months, especially in high-risk areas, can foster peaceful conflict resolution and reduce tensions.
- Public Awareness Campaigns: Launching public awareness campaigns during summer months could focus on de-escalation techniques, responsible gun ownership, and the importance of seeking help during conflicts.
- Increased Patrols in High-Risk Areas: Strategic deployment of additional police patrols or community safety teams in areas with historically higher shooting rates could deter gun violence during summer months.

**Drawbacks and Considerations**

- Data Limitations: This analysis is based on a limited time frame. A longer period and access to additional data (e.g., location details, demographics, type of shootings) could provide a more nuanced understanding.
- Reporting Bias: The data might be skewed if reporting rates differ across seasons.
- Underlying Factors: While summer weather might play a role, it's likely not the sole factor. Other potential factors might contribute like gang activity, access to firearms, or socioeconomic conditions, etc

## 3. Model: Likelihood of Cases leading to Deaths
```{r nypd2, message = FALSE, warning = FALSE}

# Calculate number of cases by borough
boro_counts <- nypd %>%
  count(BORO) %>%
  rename(CASES = n)

# Calculate number of murders by borough
murder_cases <- nypd %>%
  filter(STATISTICAL_MURDER_FLAG =="TRUE")
boro_murders <- murder_cases %>%
  count(BORO) %>%
  rename(MURDERS = n)

# Join data frames by borough
boro <- boro_counts %>%
  left_join(boro_murders, by = "BORO")

# Rename BORO to BOROUGH
boro <- boro %>%
  rename(BOROUGH = BORO)

# Calculate murder rate
boro <- boro %>%
  mutate(MURDER_RATE = round((MURDERS / CASES) * 100, 2))

# Sort by murder rate (descending)
boro <- arrange(boro, desc(MURDER_RATE))

# Round the percentages to 2 decimal places and format with % sign
boro <- boro %>%
  mutate(MURDER_RATE = sprintf("%.2f%%", MURDER_RATE))

# Print results as a table
kable(boro, align = c('l', 'l', 'l', 'l'))

# Fit linear regression model
model <- lm(MURDERS ~ CASES, data = boro)

# Summarize the model
summary(model)

# Predict values from the model
predicted <- predict(model)

# Create a data frame for plotting
plot_data <- data.frame(BOROUGH = boro$BOROUGH,
                        CASES = boro$CASES,
                        MURDERS = boro$MURDERS,
                        Predicted = predicted)

# Create plot for model
ggplot(plot_data, aes(x = CASES, y = MURDERS, label = BOROUGH)) +
  geom_point() +  # Scatter plot of actual data points
  geom_text(check_overlap = TRUE, vjust = 1.5, size = 2) +  # Add borough labels
  geom_smooth(method = "lm", se = FALSE, color = "blue") +  # Fitted regression line
  labs(x = "Number of Cases", y = "Number of Murders",
       title = "Linear Regression: Cases Leading to Murders") +
  theme_minimal()
```

### Analysis of the Model

**Model Description:**

This analysis focuses on a simple model derived from the boro data frame. We analyze the relationship between reported shootings and resulting deaths across New York City boroughs.

**Observations:**

The data suggests a positive correlation between the number of shootings and the number of murders. This implies that as shootings increase, so do fatalities.

In this context, for every one unit increase in the independent variable (shootings), there's a 0.19 unit increase in the dependent variable (murders).

Interestingly, Manhattan appears to have a slightly lower murder rate compared to other boroughs. Possible Reasons for Lower Murder Rate in Manhattan:

- Increased Police Presence: Manhattan may have a higher concentration of police officers leading to quicker response times and potentially preventing fatalities during shootings.
- Socioeconomic Factors: Manhattan has a higher average income compared to other boroughs. Lower crime rates are often associated with higher socioeconomic status.
- Medical Resources: Manhattan boasts a higher concentration of top-tier hospitals, which could improve survival rates after shootings.
- Gun Type: The type of firearms used in Manhattan shootings might be less powerful compared to other boroughs.
- Public Awareness: Manhattan residents might be more aware of personal safety measures or have quicker access to emergency response training.

**Actionable Insights:**

Law enforcement agencies in other boroughs can benchmark strategies used in Manhattan to potentially reduce fatalities. This could involve:

- Increased Patrols: Deploying additional officers in high-crime areas.
- Community Policing: Building stronger relationships between police and residents to improve trust and information sharing.
- De-escalation Training: Equipping officers with better de-escalation tactics to minimize the need to use force.

**Drawbacks and Considerations:**

- This is a simple model that only considers the outcome of shootings (death or not). It doesn't account for factors like severity of injuries, type of weapon used, or response time.
- Correlation doesn't imply causation. While the data shows a positive correlation, it doesn't necessarily mean more shootings cause more murders. Other factors might be influencing both variables.
- Generalizing from Manhattan may not be straightforward. Manhattan's unique demographics and infrastructure might not be easily replicated in other boroughs.
