---
title: "Final Project"
author: "My Le"
date: "2024-06-17"
output: github_document
---
## Background

The data examines trends during the COVID-19 pandemic from January 22, 2020, to March 9, 2023, globally and specifically in the US.

The first part of this analysis emphasizes the critical role of economic stability and healthcare investment in mitigating the global impact of COVID-19. It stresses the need for comprehensive strategies that bolster healthcare system resilience and promote broader economic well-being to effectively combat the pandemic. By addressing these interconnected aspects, countries can better prepare to navigate and mitigate the challenges posed by COVID-19, ensuring both health security and economic resilience for their populations.

The second part focuses on trends observed within the US during the pandemic. It highlights a peak in cases in 2022, which coincided with eased quarantine measures and increased social activity. This underscores the importance of maintaining social distancing measures to control transmission rates. Additionally, the analysis identifies a correlation between cases and deaths in the US, with deaths increasing over time at a slower rate compared to cases. This trend suggests that factors such as vaccination efforts and improved healthcare responses are influencing outcomes. These insights guide strategies to strengthen public health measures, enhance healthcare capacity, and mitigate both health and economic impacts of COVID-19 in the US context.

## COVID-19 Global Cases Analysis

```{r setup, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(knitr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r data1, message = FALSE}
# Retrieve the data
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv",  "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv",  "time_series_covid19_deaths_global.csv")
urls <- str_c(url_in,file_names)

# Define the data frames
global_cases <- read_csv(urls[2])
global_deaths <- read_csv(urls[4])
US_cases <- read_csv(urls[1])
US_deaths <- read_csv(urls[3])

# Tidy up Global cases
global_cases <- global_cases %>%
   rename(State = 'Province/State', Country = 'Country/Region') %>%  # Rename columns
   pivot_longer(cols = -c(State, Country, Lat, Long),
                 names_to = "date",
                 values_to = "cases") %>%
    select(-c(Lat, Long))  # Remove Lat and Long as not needed after the pivot

# Tidy up Global deaths
global_deaths <- global_deaths %>%
    rename(State = 'Province/State', Country = 'Country/Region') %>%  # Rename columns
   pivot_longer(cols = -c(State, Country, Lat, Long),
                 names_to = "date",
                 values_to = "deaths") %>%
    select(-c(Lat, Long))  # Remove Lat and Long as not needed after the pivot

# Join Global cases & deaths
global <- global_cases %>%
    full_join(global_deaths) %>%
    mutate(date = mdy(date))

# Group by "Country" and summarize total cases, deaths, deaths over cases
summary <- global %>%
  group_by(Country) %>%
  summarize(
    total_cases = sum(cases, na.rm = TRUE),
    total_deaths = sum(deaths, na.rm = TRUE)
  ) %>%
  # Add a new column 'death_rate' expressed as a percentage
  mutate(death_rate = round((total_deaths / total_cases) * 100, 2),
         death_rate_str = sprintf("%.2f%%", death_rate))  # Format as percentage string

# Sort the summary data frame by death_rate in descending order
summary_sorted <- summary %>%
  arrange(desc(death_rate))

# Select columns for kable output (excluding intermediate columns)
summary_display <- summary_sorted %>%
  select(Country, total_cases, total_deaths, death_rate_str)

# Print the data
kable(summary_display, align = 'l')
```

**Outlier Identification:**

Upon initial examination, death-to-case ratios exceeding 100% are uncommon. North Korea notably stands out as an outlier in this dataset, which may indicate potential data inaccuracies or significant underreporting of cases. This discrepancy could be attributed to challenges in identifying cases due to limited healthcare infrastructure or citizens' awareness of their health status and symptoms. Therefore, it is advisable to exclude North Korea from the dataset at this time.

```{r data2, message = FALSE}
# Filter out data for "Korea, North"
filtered_data <- summary_display %>%
  filter(Country != "Korea, North")

# Convert death_rate_str back to numeric for sorting
filtered_data$death_rate_numeric <- as.numeric(gsub("%", "", filtered_data$death_rate_str))

# Sort the data by death_rate_numeric in descending order
top10_highest <- filtered_data %>%
  arrange(desc(death_rate_numeric)) %>%
  head(10)  # Select top 10 countries with highest death rates

# Create a bar plot for top 10 highest death rates
ggplot(top10_highest, aes(x = reorder(Country, death_rate_numeric), y = death_rate_numeric)) +
  geom_bar(stat = "identity", fill = "#1f78b4") +
  labs(
    title = "Top 10 Highest Death Rates (Excluding Korea, North)",
    x = "Country",
    y = "Death Rate (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::comma_format()) +
  coord_flip()  # Flip coordinates for horizontal bars
```

### Analysis

**Observations:**

- High Death-to-Case Ratio Countries: The countries with the highest death-to-case ratios predominantly fall within the category of developing nations. These regions may experience higher mortality rates due to limited access to healthcare, lower awareness of COVID-19 prevention measures, or communal living conditions that facilitate virus transmission.
  
- Low Death-to-Case Ratio Countries: Conversely, countries with the lowest death-to-case ratios are often located in Oceania (Australia, New Zealand, Nauru, Vanuatu) or isolated geographical locations such as mountains (Bhutan) or islands (Singapore, Iceland, Marshall Islands). Factors contributing to their lower ratios may include smaller populations, geographic isolation limiting travel and virus spread, and climatic conditions potentially less favorable for virus transmission.
  
**Actionable insights:**

- Promoting Distancing Measures: Countries can learn from those with lower ratios by implementing effective distancing measures and possibly relocating households to enhance separation.

- Economic Support for Healthcare: Investing in healthcare infrastructure and emergency healthcare access can bolster a country’s ability to manage and respond to health crises, thereby improving overall resilience against the virus. Enhanced economic support can provide resources for expanding healthcare facilities, acquiring medical supplies, and ensuring healthcare workers are adequately supported.

- Healthcare Education and Campaigns: Prioritizing healthcare education and awareness campaigns can empower citizens to adopt preventive measures and strengthen public health responses.

- General Economic Enhancement: Strengthening the overall economy can indirectly contribute to better COVID-19 outcomes. Economic stability enables governments to allocate more funds towards healthcare, implement social support programs, and invest in public health education. A robust economy can also support businesses and individuals, reducing economic disparities that may exacerbate healthcare inequalities.
  
**Drawbacks:**

- Assumptions and Data Gaps: Some assumptions, such as the impact of population density or healthcare system capabilities, require further validation. Variations in reporting practices across countries can also lead to discrepancies in data accuracy.
  
- Need for Further Investigation: More comprehensive investigations into factors like population density, healthcare accessibility, and national responses to COVID-19 are necessary to provide a clearer understanding of the observed trends and variations in death-to-case ratios.
  
- Underreporting Concerns: Acknowledgment that some countries may not fully report COVID-19 cases and deaths, which can skew comparative analyses and limit actionable insights.


## COVID-19 US Cases Analysis
```{r data3, message = FALSE}

# Tidy up US cases
US_cases <- US_cases %>%
    select(-c(UID, iso2, iso3, code3, FIPS, Admin2, Combined_Key)) %>%  # Remove irrelevant columns
    rename(State = Province_State, Country = Country_Region) %>%  # Rename columns
    pivot_longer(cols = -c(State, Country, Lat, Long_),
                 names_to = "date",
                 values_to = "cases") %>%
    select(-c(Lat, Long_))  # Remove Lat and Long_ as not needed after the pivot

# Tidy up US deaths
US_deaths <- US_deaths %>%
    select(-c(UID, iso2, iso3, code3, FIPS, Admin2, Combined_Key)) %>%  # Remove irrelevant columns
    rename(State = Province_State, Country = Country_Region) %>%  # Rename columns
    pivot_longer(cols = -c(State, Country, Lat, Long_, Population),
                 names_to = "date",
                 values_to = "deaths") %>%
    select(-c(Lat, Long_))  # Remove Lat and Long_ as not needed after the pivot

# Check for duplicates in US_cases
duplicates_cases <- US_cases %>%
    group_by(State, Country, date) %>%
    filter(n() > 1)

# Check for duplicates in US_deaths
duplicates_deaths <- US_deaths %>%
    group_by(State, Country, date) %>%
    filter(n() > 1)

# Remove duplicates by summarizing cases in US_cases
US_cases <- US_cases %>%
    group_by(State, Country, date) %>%
    summarize(cases = sum(cases, na.rm = TRUE), .groups = 'drop')

# Remove duplicates by summarizing deaths and population in US_deaths
US_deaths <- US_deaths %>%
    group_by(State, Country, date) %>%
    summarize(deaths = sum(deaths, na.rm = TRUE), 
              population = first(Population), .groups = 'drop')

# Join the data frame
US <- US_cases %>%
    left_join(US_deaths, by = c("State", "Country", "date"))

# Convert date column to Date format
US <- US %>%
  mutate(date = mdy(date))

# Create the plot
ggplot(data = US, aes(x = date)) +
  geom_line(aes(y = cases, color = "Cases")) +
  geom_line(aes(y = deaths * 10, color = "Deaths")) +  # Scale deaths for better visibility
  labs(title = "COVID-19 Cases and Deaths Progression",
       x = "Date",
       y = "Cases",
       color = "Metric") +
  scale_color_manual(values = c("blue", "red"), 
                     labels = c("Cases", "Deaths")) +
  theme_minimal() +
  # Add a secondary y-axis for deaths
  scale_y_continuous(sec.axis = sec_axis(~./10, name = "Deaths", 
                                         breaks = seq(0, 250000, by = 50000))) +
  theme(axis.title.y.right = element_text(color = "red"))  # Adjust the color of the secondary y-axis label
```

### Analysis

**Observations:**

- Peak in Cases in 2022: The graph indicates a peak in the rate of COVID-19 cases in 2022, coinciding with a period when quarantine measures were relaxed and social activities resumed. This suggests that adherence to social distancing measures plays a crucial role in controlling the spread of COVID-19.

- Relationship Between Cases and Deaths: There is a clear correlation between the number of cases and deaths over time. While deaths increased throughout the year, the rate of increase slowed down over time. This trend indicates that the mortality rate relative to the number of cases decreased. Possible contributing factors include increased vaccination coverage, acquired immunity among those previously infected, and improved access to COVID-19 prevention and treatment measures.

**Actionable Insight**

- Importance of Social Distancing: The peak in cases during 2022 underscores the critical role of social distancing measures in preventing the rapid spread of COVID-19. Encouraging continued adherence to these measures during periods of high transmission can help mitigate the impact on public health.

- Vaccination and Preventive Measures: The observed decrease in the mortality rate relative to cases highlights the effectiveness of vaccination campaigns and widespread adoption of preventive measures. Promoting vaccination and ensuring access to healthcare resources remain crucial in managing COVID-19 outcomes.

**Drawbacks**

- Data Limitations: The analysis relies on aggregated data and may not capture localized variations or specific demographic factors influencing COVID-19 transmission and outcomes.

- Causal Inferences: While correlations are observed between cases, deaths, and interventions like social distancing and vaccination, establishing direct causal relationships requires more detailed epidemiological studies.

- Temporal Trends: The slowdown in death rates observed over time may also reflect changes in testing protocols, healthcare capacity, and reporting practices, which can influence the interpretation of trends.

By considering these insights and limitations, stakeholders can better formulate targeted strategies to combat COVID-19 and mitigate its impact on public health.


### Data Model

To explore the association between cases and deaths more comprehensively, we will construct a basic linear regression model.

```{r data4, message = FALSE, warning = FALSE}

# Scatter plot to visualize relationship
ggplot(US, aes(x = cases, y = deaths)) +
  geom_point() +
  labs(title = "Relationship Between COVID-19 Cases and Deaths in US",
       x = "Cases",
       y = "Deaths")

# Fit linear regression model
model <- lm(deaths ~ cases, data = US)

# Print model summary
summary(model)

# Add regression line to scatter plot
ggplot(US, aes(x = cases, y = deaths)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +  # Add linear regression line
  labs(title = "Relationship Between COVID-19 Cases and Deaths",
       x = "Cases",
       y = "Deaths")
```

**Analysis:**

- Coefficient of Cases: The coefficient of cases in the linear regression model is  *1.127e-02*. This indicates that for every unit increase in the number of COVID-19 cases, there is an associated increase of 0.01127 units in the number of deaths. In other words, for every 100 more COVID-19 cases, there is 1 more case of deaths.

- P-value: The p-value associated with the coefficient of cases is *< 2.2e-16*. This extremely low p-value suggests that the relationship between cases and deaths is statistically significant. In other words, the observed association between COVID-19 cases and deaths in the dataset is unlikely to be due to random chance.

**Interpretation:**

- Positive Relationship: The positive coefficient indicates that as the number of COVID-19 cases increases, there is a corresponding increase in the number of deaths. This aligns with the expected understanding that higher case numbers generally lead to higher mortality rates, depending on various factors including healthcare capacity, demographics, and access to treatment.

- Statistical Significance: The very low p-value indicates strong evidence against the null hypothesis (which typically assumes no relationship between cases and deaths). Therefore, we can confidently conclude that changes in COVID-19 cases are associated with changes in deaths in a statistically significant manner.

**Conclusion:**

Based on the regression analysis:

- The data supports a statistically significant positive relationship between COVID-19 cases and deaths.

- Each incremental increase in COVID-19 cases is associated with a predictable increase in deaths, as indicated by the coefficient.

These findings underscore the importance of efforts to control and manage COVID-19 transmission to mitigate its impact on mortality rates.

**Considerations:**

- Causality: While the statistical analysis shows an association, establishing causality would require additional research and consideration of other contributing factors.

- Data Quality: The accuracy and completeness of the data on COVID-19 cases and deaths are critical for the validity of the regression analysis.

- Temporal and Contextual Factors: The relationship observed may vary over time and across different regions, influenced by factors such as public health interventions, healthcare resources, and population demographics.

By understanding these statistical findings and considering the broader context, stakeholders can make informed decisions to address and mitigate the impact of COVID-19 on public health.







